Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2017-12-04 PAPER.tex
put hat's on all empirically estimated C's

\noindent where $\hat{\textbf{p}}$ is the normalized estimate of the true power spectrum. The data analyst has a choice for $
\textbf{M}$ --- for simplicity in this section we set $\textbf{M} = \textbf{I}$, although we explore other cases for the analysis 
of PAPER-64 data as explained in Section \ref{sec:Bias}.

@@

Since $\hat{\textbf{C}}$ is estimated from our data, its eigenspectrum differs from the eigenspectrum of our 
true $\textbf{C}$, and this difference has consequences on our result. An eigenspectrum ranks the eigenvalues of a matrix from 
highest to lowest and can be thought of as a spectrum of weights that are given to each spectral mode in the data. In other 
words, the eigenvalues encode the strength of different shapes in the dataset. The eigenspectrum of the identity matrix $
\textbf{I}$ is flat (all $1$'s) because it gives equal weighting to all modes. This is usually not the case for a covariance matrix, 
for which a sloped eigenspectrum means that modes are given different weights. The modes with the highest eigenvalues are 
down-weighted the most.

@@

In short, we can summarize a signal loss estimate as the ratio of $P_{\rm out}/P_{\rm in}$, evaluated at the data level $\hat{\textbf{p}}_{x}$ (our detailed computation is described in the next section). We motivate the fact that we can evaluate the output-to-input power spectrum ratio at $\hat{\textbf{p}}_{x}$ by the following reasoning.

@@

When $\textbf{R}_{r}$
is taken to be $\textbf{C}_{r}^{-1}$ and $\textbf{C}_{r}$ is estimated from the data itself, the second term of Equation \eqref{eq:pout_expand} is a cross correlation between $\textbf{C}_{r}^{-1}\textbf{x}$ and
$\textbf{C}_{r}^{-1}\textbf{e}$. As shown in \citet{switzer_et_al2015}, this cross correlation term is non-zero, and in fact negative in expectation. 
This negative cross-term power arises from a coupling between the inverse of 
$\textbf{C}_{r}$ and $\mathbf{x}$. 
Intuitively, we can see this by expanding the empirical covariance of $\textbf{r}=\textbf{x}+\textbf{e}$ in the
$|\textbf{x}|\ll|\textbf{e}|$ regime:

\begin{eqnarray}
\textbf{C}_{r} &=& \langle \textbf{rr}^{\dagger} \rangle_{t} \\
&=& \langle \textbf{xx}^{\dagger} \rangle_{t} + \langle \textbf{xe}^{\dagger} \rangle_{t} + \langle \textbf{ex}^{\dagger} \rangle_{t} + \langle 
\textbf{ee}^{\dagger} \rangle_{t},
\end{eqnarray}

\noindent where we can neglect the first term because $\textbf{x}$ is small.  Without loss of generality, we will assume
an eigenbasis of $\textbf{e}$, so that $\langle 
\textbf{ee}^{\dagger} \rangle_{t}$ is diagonal. 
The middle 
two terms, however, can have power in their off-diagonal terms due to the fact that, when averaging over a finite
ensemble, $\langle\textbf{xe}^\dagger\rangle_t$ is not zero.  As shown in Appendix C of \citet{parsons_et_al2014}, to leading order the
inversion of a matrix like $\textbf{C}_{r}$ that is diagonal-dominant (from $\langle 
\textbf{ee}^{\dagger} \rangle_{t}$) with smaller
off-diagonal terms results in a new diagonal-dominant matrix with negated off-diagonal terms. These off-diagonal
terms depend on both $\textbf{x}$ and $\textbf{e}$. When $\textbf{C}^{-1}_{r}$ is multiplied into $\textbf{x}$,
the result is a vector that is similar to $\textbf{x}$ (because of the diagonal dominance of $\textbf{C}^{-1}_{r}$), \acl{I'm not sure the previous claim follows. Why does this follow from the diagonal dominance?} but
contains a residual correlation to $\textbf{e}$ from the off-diagonal components of $\textbf{C}^{-1}$. The
correlation is negative because the product $\textbf{C}_r^{-1}\textbf{x}$ effectively squares the $\textbf{x}$-dependence
of the off-diagonal terms in $\textbf{C}^{-1}_{r}$ while retaining the negative sign that arose from the inversion
of a diagonal-dominant matrix.
% XXX [ARP: we may want to write this all out explicitly in an appendix]

The correlation between $\textbf{C}_r^{-1}\textbf{e}$ and $\textbf{C}_r^{-1}\textbf{x}$ means that we cannot compute signal loss using a signal-only 
simulation, which would yield greater values for $P_{\rm out}$ and thereby underestimate signal loss. Therefore, in our revised 
signal loss computation we use the full quantity for $P_{\rm out}$ as defined in Equation \eqref{eq:sigloss}, which subtracts the 
weighted power spectrum of the data from the weighted power spectrum of data plus EoR.

We thus see from this subsection that signal loss generically arises because of couplings between $\textbf{C}^{-1}_{r}$ and $\mathbf{x}$. In principle, using $\textbf{C}^{-1}_{r}$ in place of the true inverse covariance $\textbf{C}^{-1}$ provides yet another source of multiplicative bias, which results in a modification to the treatment shown here. Such a modification is accounted for in the demonstration of Appendix \ref{sec:sigloss_appendix}, where we additionally discard the assumption that $|\textbf{x}|\ll|\textbf{e}|$. However, despite the loosening of these approximations, one sees that the qualitative message (of anti-correlations resulting in signal loss) remains intact.

@@

\begin{eqnarray}
P_{\rm in,\alpha} &=&  M_{\alpha}\textbf{e}^{\dagger}\textbf{I}\textbf{Q}^{\alpha}\textbf{I}\textbf{e} \\
P_{\rm out,\alpha} &=&  M_{\alpha}\textbf{r}^{\dagger}\textbf{C}_{r}^{-1}\textbf{Q}^{\alpha}\textbf{C}_{r}^{-1}\textbf{r} \nonumber \\
&&\phantom{=====} 
-  M_{\alpha}\textbf{x}^{\dagger}\textbf{C}_{x}^{-1}\textbf{Q}^{\alpha}\textbf{C}_{x}^{-1}\textbf{x} 
\end{eqnarray}

@@

\begin{itemize}
\item Substantial signal loss can result when weighting data using empirically estimated covariances (Section 
\ref{sec:SiglossOverview}). Loss of the $21$ cm signal is especially significant if fewer modes
exist in the data. Hence, there exists a trade-off between sensitivity driven 
time-averaging techniques such as fringe-rate filtering and signal loss. 
\item Signal injection and recovery simulations can be used to quantify signal loss (Section \ref{sec:Sigloss}). However, a 
signal-only simulation (i.e. comparing an unweighted vs. weighted power spectrum of EoR only) can underestimate loss by 
failing to account for spurious correlations between the data and signal.
\item Errors that are estimated via bootstrapping can be underestimated if samples in the dataset are significantly correlated 
(Section \ref{sec:ErrorOverview}). If the number of independent samples in a dataset is well-determined, bootstrapping is a 
simple and accurate way of estimating errors.
\item Meaningful null tests are vital to validate an EoR detection (Section \ref{sec:JackknifeOverview}). Similarly, performing 
jackknife tests along multiple axes of a dataset is necessary for confidence in an EoR detection and can also be used to tease 
out systematics.
\end{itemize}

>>>>

\noindent where $\hat{\textbf{p}}$ is the normalized estimate of the true power spectrum. The data analyst has a choice for $
\textbf{M}$ --- for simplicity in this section we choose $\textbf{M}$ to be diagonal, although we explore other cases for the analysis 
of PAPER-64 data as explained in Section \ref{sec:Bias}.

@@

If the covariance is estimated from our data, its eigenspectrum differs from the eigenspectrum of the true covariance $\textbf{C}$, and this difference has consequences on our result. An eigenspectrum ranks the eigenvalues of a matrix from 
highest to lowest and can be thought of as a spectrum of weights that are given to each spectral mode in the data. In other 
words, the eigenvalues encode the strength of different shapes in the dataset. The eigenspectrum of the identity matrix $
\textbf{I}$ is flat (all $1$'s) because it gives equal weighting to all modes. This is usually not the case for a covariance matrix, 
for which a sloped eigenspectrum means that modes are given different weights. The modes with the highest eigenvalues are 
down-weighted the most.

@@

In short, we can summarize a signal loss estimate as the ratio of $P_{\rm out}/P_{\rm in}$, evaluated at the data level $\hat{\textbf{p}}_{x}$. We motivate the fact that we can evaluate the output-to-input power spectrum ratio at $\hat{\textbf{p}}_{x}$ by the following reasoning.

@@

When $\textbf{R}_{r}$
is taken to be $\hat{\textbf{C}}_{r}^{-1}$, the second term of Equation \eqref{eq:pout_expand} is a cross-correlation between $\hat{\textbf{C}}_{r}^{-1}\textbf{x}$ and
$\hat{\textbf{C}_{r}}^{-1}\textbf{e}$. As shown in \citet{switzer_et_al2015}, this cross-correlation term is non-zero, and in fact negative in expectation. 
This negative cross-term power arises from a coupling between the inverse of 
$\hat{\textbf{C}}_{r}$ and $\mathbf{x}$. 
Intuitively, we can see this by expanding the empirical covariance of $\textbf{r}=\textbf{x}+\textbf{e}$ in the
$|\textbf{x}|\ll|\textbf{e}|$ regime:

\begin{eqnarray}
\hat{\textbf{C}}_{r} &=& \langle \textbf{rr}^{\dagger} \rangle_{t} \nonumber \\ 
&=& \langle \textbf{xx}^{\dagger} \rangle_{t} + \langle \textbf{xe}^{\dagger} \rangle_{t} + \langle \textbf{ex}^{\dagger} \rangle_{t} + \langle 
\textbf{ee}^{\dagger} \rangle_{t},
\end{eqnarray}

\noindent where we can neglect the first term because $\textbf{x}$ is small.  Without loss of generality, we will assume
an eigenbasis of $\textbf{e}$, so that $\langle 
\textbf{ee}^{\dagger} \rangle_{t}$ is diagonal. 
The middle 
two terms, however, can have power in their off-diagonal terms due to the fact that, when averaging over a finite
ensemble, $\langle\textbf{xe}^\dagger\rangle_t$ is not zero.  As shown in Appendix C of \citet{parsons_et_al2014}, to leading order the
inversion of a matrix like $\hat{\textbf{C}}_{r}$ that is diagonal-dominant (from $\langle 
\textbf{ee}^{\dagger} \rangle_{t}$) with smaller
off-diagonal terms results in a new diagonal-dominant matrix with negated off-diagonal terms. These off-diagonal
terms depend on both $\textbf{x}$ and $\textbf{e}$. When $\hat{\textbf{C}}^{-1}_{r}$ is multiplied into $\textbf{x}$,
the result is a vector that is similar to $\textbf{x}$ but
contains a residual correlation to $\textbf{e}$ from the off-diagonal components of $\hat{\textbf{C}}^{-1}_{r}$. The
correlation is negative because the product $\hat{\textbf{C}}_r^{-1}\textbf{x}$ effectively squares the $\textbf{x}$-dependence
of the off-diagonal terms in $\hat{\textbf{C}}^{-1}_{r}$ while retaining the negative sign that arose from the inversion
of a diagonal-dominant matrix.
% XXX [ARP: we may want to write this all out explicitly in an appendix]

The correlation between $\hat{\textbf{C}}_r^{-1}\textbf{e}$ and $\hat{\textbf{C}}_r^{-1}\textbf{x}$ means that we cannot compute signal loss using a signal-only 
simulation, which would yield greater values for $P_{\rm out}$ and thereby underestimate signal loss. Therefore, in our revised 
signal loss computation we use the full quantity for $P_{\rm out}$ as defined in Equation \eqref{eq:sigloss}, which subtracts the 
weighted power spectrum of the data from the weighted power spectrum of data plus EoR.

We thus see from this subsection that signal loss generically arises because of couplings between $\hat{\textbf{C}}^{-1}_{r}$ and $\mathbf{x}$. In principle, using $\hat{\textbf{C}}^{-1}_{r}$ in place of the true inverse covariance $\textbf{C}^{-1}$ provides yet another source of multiplicative bias, which results in a modification to the treatment shown here. Such a modification is accounted for in the demonstration of Appendix \ref{sec:sigloss_appendix}, where we additionally discard the assumption that $|\textbf{x}|\ll|\textbf{e}|$. However, despite the loosening of these approximations, one sees that the qualitative message (of anti-correlations resulting in signal loss) remains intact.

@@

\begin{eqnarray}
P_{\rm in,\alpha} &=&  M_{\alpha}\textbf{e}^{\dagger}\textbf{I}\textbf{Q}^{\alpha}\textbf{I}\textbf{e} \\
P_{\rm out,\alpha} &=&  M_{\alpha}\textbf{r}^{\dagger}\hat{\textbf{C}}_{r}^{-1}\textbf{Q}^{\alpha}\hat{\textbf{C}}_{r}^{-1}\textbf{r} \nonumber \\
&&\phantom{=====} 
-  M_{\alpha}\textbf{x}^{\dagger}\hat{\textbf{C}}_{x}^{-1}\textbf{Q}^{\alpha}\hat{\textbf{C}}_{x}^{-1}\textbf{x} 
\end{eqnarray}

@@

\begin{itemize}
\item Substantial signal loss can result when weighting data using empirically estimated covariances (Section 
\ref{sec:SiglossOverview}). Loss of the $21$ cm signal is especially significant if fewer modes
exist in the data. Hence, there exists a trade-off between sensitivity driven 
time-averaging techniques such as fringe-rate filtering and signal loss. 
\item Signal injection and recovery simulations can be used to quantify signal loss (Section \ref{sec:Sigloss}). However, a 
signal-only simulation (i.e. comparing an unweighted vs. weighted power spectrum of EoR only) can underestimate loss by 
failing to account for correlations between the data and signal (Section \ref{sec:Illustration}).
\item Errors that are estimated via bootstrapping can be underestimated if samples in the dataset are significantly correlated 
(Section \ref{sec:ErrorOverview}). If the number of independent samples in a dataset is well-determined, bootstrapping is a 
simple and accurate way of estimating errors.
\item Meaningful null tests are vital to validate an EoR detection (Section \ref{sec:JackknifeOverview}). Similarly, performing 
jackknife tests along multiple axes of a dataset is necessary for confidence in an EoR detection and can also be used to tease 
out systematics.
\end{itemize}
